A performance assessment of evolutionary algorithms in a volunteer computing system

### 2016-03-20 PPSN2016/paper.Rnw
Back to 11 pages after fixes

graph2 & 1173561.5 & 6 & 4 & 761.5 & 22.0 & 0.0520509 & 36 & 73 & 129.04545\\
\hline
reboot & 796214.0 & 9 & 6 & 646.0 & 67.0 & 0.0955121 & 45 & 168 & 75.88889\\
\hline
cachereboot & 392917.0 & 12 & 12 & 576.0 & 79.0 & 0.1575092 & 55 & 225 & 43.28571\\
\hline
\end{tabular}
%% \end{tabular}
%% Initially generated with this code, 
% <<experiments, cache=FALSE,echo=FALSE>>=
%experiment.aggregates$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%kable(experiment.aggregates)
%@ %def 
\end{center}
\end{table}
%
%
%
Since the point of using volunteer computing system is to cut costs,
all experiments were set up using the OpenShift
PaaS, which provides a free tier, making the whole experiment cost
equal to \$0.00. In fact, the NodIO system can be deployed to any
system as long as it can run node.js and has access to the filesystem, for
log files, if available. All these logs are available upon
request. Processed files are available in the same repository that
hosts this paper, as well as the scripts needed to process them.

%
%
\begin{figure}[htbp]
\centering
<<zipf, cache=FALSE,echo=FALSE,fig.height=3>>=
ggplot(ips.data,aes(x=idx,y=PUTs,group=Experiment,color=Experiment))+geom_point()+scale_y_log10()+ scale_colour_discrete(labels=c('base','graph1','graph2','reboot','cachereboot'))+scale_x_log10()
@ %def graph for IPs and puts
%
\caption{Ranked and normalized number of contributions (PUTs) per IP,
  plotted in a log $y$ scale.\label{fig:zipf}}
\end{figure}
%
Every run had slightly different conditions, although in some cases it
was just a server reboot and a new round of announcements. Each run
batch included at least 30 completions, that is, evolution until the
solution was found. The number of runs was controlled by polling an
URL that indicated the number of runs made in the present batch so
far. Every volunteer contributed a certain number of individuals
(represented by the PUT HTTP request) to the server. The ranked number
of contributions per unique IP is shown in Figure \ref{fig:zipf} in a
log-log scale. Besides following the usual Zipf's law found in our
previous work and having a very similar appearance for all 5 sets, it
is interesting to see how some sets have a number of contributions per
IP that are slightly superior to the rest, specially so in {\tt
  graph2}. The last two sets, {\tt reboot} and {\tt cachereboot}, also
show a regime in the middle of the graph that shows that some users
are contributing more than in other similar sets. In general, these
differences in behavior are due more to the specific users that show
up in these experiments than to general laws, but it should also be
noted that those behaviors can be encouraged by design.

A summary of the experiments and  their results is shown in Table
\ref{tab:runs}. This table shows the median time needed to find the
solution as well as the median number of IPs and PUTs. In this case we
distinguish between the total number of clients participating in every
particular run, and the {\em actual} number whose contributions were
accepted into the cache, which was set up to accept those only if {\em
  new} chromosomes were sent to the server in order to avoid overlaps
between one run and the next. That is why, in both cases, the
``Contributing'' column includes a value that is less than the total
number. The final column includes the total number of runs in that
particular batch, finished roughly after a minimum number of 30 was reached.

The baseline, called {\em base} in the Table, uses the
mechanism of dropping contributions from clients if they are repeated
to avoid overlap. Clients still get a random chromosome from the
server, so they can in fact proceed with the algorithm and even finish
it. The fact that these users were not contributing to the pool was
conveyed by inserting a graph in the client that represented the size
of the cache, labeled ``How am I contributing''. The user could
then realize if his contribution was being 0 and thus do something
about it; these are labeled {\tt graph1} and 2. However, a stuck
client could still add to the pool if evolution got unstuck, so we
added a mechanism for rebooting the client, that is, reloading the
page, if the cache size was found to be less than 1, which indicated a
run recently initiated. This batch of runs is labelled {\tt
  reboot}. However, we found that this size of the cache was actually
skipped by some slow clients or too fast evolutions, so that in the
last case, labeled {\tt cachecrash}, clients rebooted if the cache
collapsed by more than a certain length, indicating a cache that had
been voided and was filling again. Our objective was twofold: to
encourage engagement by the users and also to increase the involvement
of every client in the common experiment, eliminating at the same time
overlaps between runs.

%
\begin{table}
  \caption{Experiments with {\em overlapped} runs filtered out: summary of results. \label{tab:filtered}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Experiment & Median Time & \multicolumn{2}{|c|}{Median IPs} &\multicolumn{2}{|c|}{Median PUTs} & \#Runs &  Puts/IP \\
&   (ms)        & Total & Used & Total & Used &  & \\
\hline
base & 293685 & 13.0 & 8.0 & 483.0 & 57.0 & 47 & 35.50000\\
\hline
graph1 & 768053 & 12.0 & 5.0 & 601.0 & 79.0 & 21 & 45.84615\\
graph2 & 1477969 & 6.0 & 4.0 & 901.0 & 52.0 & 25 & 149.42857\\
\hline
reboot & 799471 & 9.5 & 6.5 & 656.5 & 67.0 & 44 & 76.48016\\
\hline
cachereboot & 463347 & 12.0 & 12.0 & 638.5 & 80.5 & 52 & 44.25911\\
\hline
\end{tabular}
%% Initially generated with this code, 
%%<<experiments.filtered, cache=FALSE,echo=FALSE>>=
%%experiment.aggregates.filtered$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%%kable(experiment.aggregates.filtered)
%%@ %def 
%
\end{center}
\end{table}
%
So we will have to check whether we have effectively eliminated these
overlaps. In order to do that, we will have to establish a threshold
under which we can reasonably expect that the solution cannot be found. In every
evolutionary algorithm it might happen, by chance, that the solution
is found in just a few evaluations, but that is usually not the
case. By looking at the logs in the last case we have established that
minimum as 16 PUTs; in less than 1600 generations no solution to that
difficult problem will be found. Let us filter then the runs,
eliminating those that used less than 1600 generations to find the
solution which can reasonably be said that they did so because some
browser ``carried over'' a chromosome from the previous run. The
results are shown in Table \ref{tab:filtered}. The first change is in
the column \#Runs, that shows that most of the runs in the initial
configurations found the solution thanks to these carried over
clients. However, in the last two rows, the ones that rebooted clients
when they detected changes in cache, the difference is minimal. This
means that one objective has been reached: if clients reboot when they
detect changes in the cache that imply a new run, overlaps are almost
eliminated. Going back to Table \ref{tab:runs}, we can see than the
ratio of effective contributions, those that feed the cache, increases
from around 13\% in the first case to almost 16\% in the last,
reaching lows of less than 6\% in some cases ({\tt graph2}). This
indicates that rebooting the clients has the positive effect of
increasing the contribution of clients to the common pool. However, we
should remark that the clients will still stop contributing if they
get stuck in a particular solution, so the ratio keeps being low
resulting in a slow change in the cache and a lower diversity of
clients that tap that cache. At any rate, an increased number of
effective PUTs will have a positive impact on the algorithm.

>>>>

graph2 & 1173561.5 & 6 & 4 & 761.5 & 22.0 & 0.0520509 & 36 & 73 & 129.04545\\
\hline
reboot & 796214.0 & 9 & 6 & 646.0 & 67.0 & 0.0955121 & 45 & 168 & 75.88889\\
\hline
cachereboot & 392917.0 & 12 & 12 & 576.0 & 79.0 & 0.1575092 & 55 & 225 & 43.28571\\
\hline
\end{tabular}
%% \end{tabular}
%% Initially generated with this code, 
% <<experiments, cache=FALSE,echo=FALSE>>=
%experiment.aggregates$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%kable(experiment.aggregates)
%@ %def 
\end{center}
\end{table}
%
%
%
Since the point of using volunteer computing system is to cut costs,
all experiments were set up using the OpenShift
PaaS, which provides a free tier, making the whole experiment cost
equal to \$0.00. In fact, the NodIO system can be deployed to any
system as long as it can run node.js and has access to the filesystem, for
log files, if available%% . All these logs are available upon
%% request. Processed files are available in the same repository that
%% hosts this paper, as well as the scripts needed to process them.

%
%
\begin{figure}[htbp]
\centering
<<zipf, cache=FALSE,echo=FALSE,fig.height=3>>=
ggplot(ips.data,aes(x=idx,y=PUTs,group=Experiment,color=Experiment))+geom_point()+scale_y_log10()+ scale_colour_discrete(labels=c('base','graph1','graph2','reboot','cachereboot'))+scale_x_log10()
@ %def graph for IPs and puts
%
\caption{Ranked and normalized number of contributions (PUTs) per IP,
  plotted in a log $y$ scale.\label{fig:zipf}}
\end{figure}
%
Every run had slightly different conditions, although in some cases it
was just a server reboot and a new round of announcements. Each run
batch included at least 30 completions, that is, evolution until the
solution was found. The number of runs was controlled by polling an
URL that indicated the number of runs made in the present batch so
far. Every volunteer contributed a certain number of individuals
(represented by the PUT HTTP request) to the server. The ranked number
of contributions per unique IP is shown in Figure \ref{fig:zipf} in a
log-log scale. Besides following the usual Zipf's law found in our
previous work and thus having a very similar appearance for all 5 sets, it
is interesting to see how some sets have a number of contributions per
IP that are slightly superior to the rest, specially so in {\tt
  graph2}. The last two sets, {\tt reboot} and {\tt cachereboot}, also
show a regime in the middle of the graph that shows that some users
are contributing more than in other similar sets. In general, these
differences in behavior are due more to the specific users that show
up in these experiments than to general laws, but it should also be
noted that those behaviors can be encouraged by design.

A summary of the experiments and  their results is shown in Table
\ref{tab:runs}. This table shows the median time needed to find the
solution as well as the median number of IPs and PUTs. In this case we
distinguish between the total number of clients participating in every
particular run, and the {\em actual} number whose contributions were
accepted into the cache, which was set up to accept those only if {\em
  new} chromosomes were sent to the server in order to avoid overlaps
between one run and the next. That is why, in both cases, the
``Contributing'' column includes a value that is less than the total
number. The next column includes the total number of runs in that
particular batch, finished roughly after a minimum number of 30 was reached.

The baseline, called {\em base} in the Table, uses the
mechanism of dropping contributions from clients if they are repeated
to avoid overlap. Clients still get a random chromosome from the
server, so they can in fact proceed with the algorithm and even finish
it. However, this was not known to the volunteer, so the fact that they were not contributing to the pool was
conveyed by inserting a graph in the client that represented the size
of the cache, labeled ``How am I contributing''. The user could
then realize there was no contribution and do something
about it. These sets of runs are labeled {\tt graph1} and 2. However, a stuck
client could still add to the pool if it found a new optimum, so we
added a mechanism for rebooting the client, that is, reloading the
page, if the cache size was found to be less than 1, which indicated a
run recently started. This batch of runs is labelled {\tt
  reboot}. This size of the cache was not really noticed by some slow clients or too fast runs, so that in the
last case, labeled {\tt cachecrash}, clients rebooted if the cache
collapsed by more than a certain length, indicating a cache that had
been voided and was filling again. Our objective was twofold: to
encourage engagement by the users and also to increase the involvement
of every client in the common experiment, eliminating at the same time
overlaps between runs.

%
\begin{table}
  \caption{Experiments with {\em overlapped} runs filtered out: summary of results. \label{tab:filtered}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Experiment & Median Time & \multicolumn{2}{|c|}{Median IPs} &\multicolumn{2}{|c|}{Median PUTs} & \#Runs &  Puts/IP \\
&   (ms)        & Total & Used & Total & Used &  & \\
\hline
base & 293685 & 13.0 & 8.0 & 483.0 & 57.0 & 47 & 35.50000\\
\hline
graph1 & 768053 & 12.0 & 5.0 & 601.0 & 79.0 & 21 & 45.84615\\
graph2 & 1477969 & 6.0 & 4.0 & 901.0 & 52.0 & 25 & 149.42857\\
\hline
reboot & 799471 & 9.5 & 6.5 & 656.5 & 67.0 & 44 & 76.48016\\
\hline
cachereboot & 463347 & 12.0 & 12.0 & 638.5 & 80.5 & 52 & 44.25911\\
\hline
\end{tabular}
%% Initially generated with this code, 
%%<<experiments.filtered, cache=FALSE,echo=FALSE>>=
%%experiment.aggregates.filtered$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%%kable(experiment.aggregates.filtered)
%%@ %def 
%
\end{center}
\end{table}
%
The first thin we will have to check is whether we have effectively eliminated these
overlaps. In order to do that, we will have to establish a threshold
under which we can reasonably expect that the solution cannot be found. In every
evolutionary algorithm it might happen, by chance, that the solution
is found in just a few evaluations, but that is usually not the
case. By looking at the logs in the last case we have established that
minimum as 16 PUTs; in less than 1600 generations no solution to that
difficult problem will be found. Let us filter then the runs,
eliminating those that used less than 1600 generations to find the
solution which can reasonably be said that they did so because some
browser ``carried over'' a chromosome from the previous run. The
results are shown in Table \ref{tab:filtered}. The first change is in
the column \#Runs, that shows that most of the runs in the initial
configurations found the solution thanks to these carried over
clients. However, in the last two rows, the ones that rebooted clients
when they detected changes in cache, the difference is minimal. This
means that one objective has been reached: if clients reboot when they
detect changes in the cache that imply a new run, overlaps are almost
eliminated. Going back to Table \ref{tab:runs}, we can see than the
ratio of effective contributions, those that feed the cache, increases
from around 13\% in the first case to almost 16\% in the last,
reaching lows of less than 6\% in some cases ({\tt graph2}). This
indicates that rebooting the clients has the positive effect of
increasing the contribution of clients to the common pool. However, we
should remark that the clients will still stop contributing if they
get stuck in a particular solution, so the ratio keeps being low
resulting in a slow change in the cache and a lower diversity of
clients that tap that cache. At any rate, an increased number of
effective PUTs will have a positive impact on the algorithm.
